The evolution of gene expression in the term placenta of viviparous mammals

### 2015-12-18 manuscript/manuscript.Rnw
clean up sequencing discussion

Fetal tissue was collected from the placenta of six mammalian species
(\human [human], \bonobo [bonobo], \dog [domestic dog], \cow [domestic
cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \gallili [?], \carmelli [?]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer’s instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and standard Illumina protocol. Paired-end sequencing was
performed using the Illumina Genome Analyzer II with an insert size of
230 base pairs (bp). Read lengths for species are shown in
\cref{tab:ref_genome_aligned_species}.

@@

For more complete comparisons additional sequences of placenta in
\mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\DLA{citation}, \sheep \DLA{citation}, and \boar \DLA{citation} were
obtained from the \ac{SRA} (Supplemental \cref{tab:srr_numbers}).

@@

\Cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
All reads were submitted to the NCBI Sequence Read Archive (SRA)
(Accession Number XXX-YYY).

\DLA{Need to do an embargoed submission to SRA}

The percentage of bases that mapped to Ensembl annotated mRNA ranged
from ~35\% in elephant to ~85\% in human, with percentages of bases
mapping to annotated protein-coding segments of genes ranging from
~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
ranged from ~43\% in elephant to 80\% in bonobo. The number of
annotated transcripts to which reads mapped in each respective genomes
ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
greater degree of mapping to annotated protein-coding sequences in
humans corresponds to the greater number of Ensembl annotated
sequences in the human genome compared to the other five reference
genomes (Flicek, 2012).

\DLA{Summarize all of this in a table}

The top 50 expressed genes in each genome analyzed are shown in
\DLA{Tables 2-7}, and all genes for each transcriptome are listed in
\DLA{Supplementary Tables X-XX} [These are large Excel files to be
included with the Supplemental Material]. In addition, specific genes
including placental hormones and hormone-related genes, galectins, and
insulin-like growth factor related genes are shown with \ac{FPKM} values
side to side for all transcriptomes including the reanalysis of mouse
from \cite{Wang.ea2011:surveyfornovelimprinted} in \DLA{Table 8}.
Below we describe each transcriptome separately, specifically
concentrating on highly-expressed genes.

>>>>

Fetal tissue was collected from the placenta of six mammalian species
(\human [human], \bonobo [bonobo], \dog [domestic dog], \cow [domestic
cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \gallili [?], \carmelli [?]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and the standard Illumina protocol. Paired-end sequencing
was performed using the Illumina Genome Analyzer II with an insert
size of 230 base pairs (bp). Read lengths for species are shown in
\cref{tab:ref_genome_aligned_species}.

@@

For more complete evolutionary comparisons additional sequences of
placenta in \mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\DLA{citation; e-mailed Alec Archibald from the ISGC about what the
  right one should be}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
\ac{SRA} (Supplemental \cref{tab:srr_numbers}).

@@

\Cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (cref{tab:ref_genome_aligned_species}).
The top 50 expressed genes in each genome analyzed are shown in
\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elphant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
and all genes in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

% The percentage of bases that mapped to Ensembl annotated mRNA ranged
% from ~35\% in elephant to ~85\% in human, with percentages of bases
% mapping to annotated protein-coding segments of genes ranging from
% ~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
% from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
% from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
% ranged from ~43\% in elephant to 80\% in bonobo. The number of
% annotated transcripts to which reads mapped in each respective genomes
% ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
% and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
% greater degree of mapping to annotated protein-coding sequences in
% humans corresponds to the greater number of Ensembl annotated
% sequences in the human genome compared to the other five reference
% genomes (Flicek, 2012).
